Refactoring the `nls()` function in R

### 2022-06-20 RefactoringNLS/RefactoringNLS.Rmd
a few changes

In the manual page for `nls()` in R 4.0.0 there is the warning

@@

We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
Unfortunately, the meaning of "iteration" varies among programs and requires 
careful examination of the
code. We prefer to use the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) when these limits
are reached. Generally we prefer larger limits than the default `maxiter=50` of `nls()`,
but that may reflect the more difficult problems we have encountered because 
users consult us when standard tools have given unsatisfactory results.

@@

gets over 4000 hits that are spread over some years. This could be because the 
Jacobian is poorly approximated (see **Jacobian computation** below). However,
it is common in nonlinear least squares computations that the Jacobian is very close
to singular for some values of the model parameters. In such cases we need to find
an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
In our work we prepared experimental prototypes of `nls` that incorporate
stabilizations, but integration with all the features of `nls()` has proved difficult.

@@

```{r SESingGrad}
reala = -3; realb = 5; realc = 0.5; realr = 0.7; realm = 1 # underlying parameters
x = 1:11 # x values; 11 data points
y=reala + realb * realr^(x - realm) + realc * x # linear+exponential function
testdat = data.frame(x, y) # save in a data frame
strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
jform <- y ~ a + b * r^(x - m) + c * x # Formula
library(nlsr)
linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
linexp2 # Note singular values of Jacobian in rightmost column
# We get an error with nls()
linexp=try(nls(jform, data = testdat, start = strt, trace = F))
```

@@

The typical default action, which can be seen by using the command `getOption("na.action")`
is `na.omit`. This option essentially omits from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much the same for computations, but keeps the rows with NA elements so 
that predictions are in the correct row position. We recommend that workers actually test 
output to verify behaviour is as wanted.

@@

`model` is an argument to the `nls()` function, which is documented as

@@

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do NOT like this approach, but it is "the R way". R allows users to leave many 
objects in the default (.GlobalEnv) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit, and any such file in the path 
when R is launched will be loaded. The overwhelming proportion of R users in our
acquaintance avoid saving the workspace because of the danger of lurking data and
functions which may cause unwanted results.

@@

Other concerns are revealed by the example below. Here the `SSlogis` selfStart function 
is used to generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $y \sim Asym/(1+exp((xmid-tt)/scal))$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $y \sim b1/(1 + b2*exp(-b3*t))$.

The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
<!-- ``` -->
<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
<!--     selfStart( -->
<!--         function(input, Asym, xmid, scal) -->
<!--         { -->
<!--               .expr1 <- xmid - input -->
<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
<!--               .expr4 <- 1 + .expr3 -->
<!--               .value <- Asym/.expr4 -->
<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
<!--               { -->
<!--             		  .expr10 <- .expr4^2 -->
<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
<!--                   .grad[, "Asym"] <- 1/.expr4 -->
<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
<!-- 		              .grad[, "scal"] <- xm * .e2 -->
<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
<!--                   attr(.value, "gradient") <- .grad -->
<!--               } -->
<!--               .value -->
<!--         }, -->
<!--         initial = function(mCall, data, LHS, ...) { -->
<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
<!--               if(nrow(xy) < 4) { -->
<!--                   stop("too few distinct input values to fit a logistic model") -->
<!--               } -->
<!--               z <- xy[["y"]] -->
<!--               ## transform to proportion, i.e. in (0,1) : -->
<!--               rng <- range(z); dz <- diff(rng) -->
<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
<!--               aux <- coef(lm(x ~ z, xy)) -->
<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
<!--                                data = xy, -->
<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
<!--                                algorithm = "plinear", ...)) -->
<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
<!--                        mCall[c("Asym", "xmid", "scal")]) -->
<!--         }, -->
<!--         parameters = c("Asym", "xmid", "scal")) -->
<!-- ``` -->
This R function includes analytic expressions for the Jacobian ("gradient").
These could be be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example model $y \sim a*exp(-b*x) + c*exp(-d*x)$
would avoid a singular Jacobiam because $b$ and $d$ were equal in value. 
A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
could be provided quite simply instead of all 1's.

@@

```
A list of:

m	        an nlsModel object incorporating the model.
data  	    the expression that was passed to nls as the data argument. The actual data values 
              are present in the environment of the m components, e.g., environment(m$conv).
call	    the matched call with several components, notably algorithm.
na.action 	the "na.action" attribute (if any) of the model frame.
dataClasses	the "dataClasses" attribute (if any) of the "terms" attribute of the model frame.
model	    if model = TRUE, the model frame.
weights	    if weights is supplied, the weights.
convInfo	a list with convergence information.
control	    the control list used, see the control argument.

There are also two deprecated items if algorithm = "port" fit only. These are
convergence (a code = 0 for convergence) and message. These are available from convInfo.
```

@@

The `data` return element is an R symbol. To actually access the data from this 
element, we need to use the syntax

@@

However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using

@@

```
coefficients    A named vector giving the parameter values at the supposed solution.
ssquares        The sum of squared residuals at this set of parameters.
resid           The residual vector at the returned parameters.
jacobian        The jacobian matrix (partial derivatives of residuals w.r.t. the 
                parameters) at the returned parameters.
feval           The number of residual evaluations (sum of squares computations) used.
jeval           The number of Jacobian evaluations used.
```

@@

```{r tetrarun}
time=c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
conc = c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
NLSdata <- data.frame(time,conc)
NLSstart <-c(lrc1=-2,lrc2=0.25,A1=150,A2=50) # a starting vector (named!)
NLSformula <-conc ~ A1*exp(-exp(lrc1)*time)+A2*exp(-exp(lrc2)*time)
tryit <- try(nls(NLSformula, data=NLSdata, start=NLSstart, trace=TRUE))
print(tryit)
```

@@

One obvious case is that nonlinear modeling tools are a superset of linear ones.
Yet the explicit model `y ~ a*x + b` does not work with the linear modeling 
function `lm()`, which requires this model to be specified as `y ~ x`.

Within `nls()`, consider the 
following FOUR different calling sequences for the same problem, though an intuitive choice,
labelled fm2a, will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing we do
not foresee a resolution for this issue. In the example we have NOT evaluated
the commands to save space.

@@

## using conditional linearity - leave out the Asym parameter
fm2 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1, start = list(xmid = 0, scal = 1),
                 algorithm = "plinear")
summary(fm2)

## without conditional linearity
fm3 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1,
                 start = list(Asym = 3, xmid = 0, scal = 1))
summary(fm3)

## using Port's nl2sol algorithm
fm4 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
                 algorithm = "port"))
summary(fm4)

## using conditional linearity AND Asym does NOT work
fm2a <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
                 algorithm = "plinear", trace = TRUE))
summary(fm2a)
```

@@

For `nls()` the concern is demonstrated by a short email query from John Nash
to Doug Bates and his reply.
This is NOT a criticism of Prof. Bates (or any other) work, but a reflection on 
how difficult it is to develop code in this subject area and to keep it maintainable. 
We have experienced similar loss of understanding for some of our own codes.

@@

> The algorithm = "port" code appears unfinished, and does not even check that the starting 
  value is within the bounds. Use with caution, especially where bounds are supplied.

@@

```{r chkbound}
# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
wf <- data.frame(y=weed, tt=tt)
st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt)) ## Scaled model
require(minpack.lm)
require(nlsr)
# Hobbs scaled problem with bounds, formula specification
anlxb1 <- nlxb(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
cat("Using nlsr::nlxb():")
print(anlxb1)
## nlsLM seems NOT to work properly with bounds
anlsLM1b <- nlsLM(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
cat("\n"); cat("using minpack.lm::nlsLM():")
print(anlsLM1b)
anlsb<-nls(wmods, start=st, data=wf, algorithm="port", lower=c(0,0,0), upper=c(2,6,3))
cat("Using nls with 'port':")
print(anlsb)
```

@@

Bounds on parameters raise some interesting and difficult questions about how uncertainty
in parameter estimates should be computed or reported. That is, the traditional "standard
errors" are generally taken to imply symmetric intervals about the point estimate in which
the parameter may be expected to be found with some probability under certain assumptions.
Bounds change those assumptions. Hence `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.

@@

We want

@@

- to streamline the overall `nls()` infrastructure. By this we
  mean a re-factoring the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

@@

While not consistent with the `lm()` specification of linear models, 
nonlinear models need all parameters specified. 
That is, the model formula should be written down as it would be computed.
Thus a linear model (which is valid input for nonlinear estimatin) should
be specified `y ~ a*x + b` rather than `y ~ x`. 
Moreover, we want to avoid the partially linear parameters appearing 
in the result object as `.lin1`, `.lin2`, etc., where the actual position 
of parameters in the model is not explicit.

@@

A further complication of model specification in R is that some formulas require
the user to surround a term with `I(  )` to inhibit interpretation of arithmetic 
operators. We would prefer that formulas be usable without this complication.

@@

Indexed models clearly have a place in some areas of research. 
<!-- We need to be able to process models, as the example above, where the -->
<!-- model formula is  -->
We need to be able to process models such as 
`Length ~ a[Strip] + b[Strip] * exp(-Conc/th)` where `Strip` is the index. Data
`Length` and `Conc` are available for all observations and parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` depend on the index, which
could be an experimental run number, or some other categorization.

>>>>

In the manual page for `nls()` in R 4.0.0 there is the warning:

@@

We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
Unfortunately, the meaning of "iteration" varies among programs and requires 
careful examination of the
code. We prefer to use the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) when these limits
are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
but that may reflect the more difficult problems we have encountered because 
users consult us when standard tools have given unsatisfactory results.

@@

gets over 4000 hits that are spread over some years. This could be because the 
Jacobian is poorly approximated (see **Jacobian computation** below). However,
it is common in nonlinear least squares computations that the Jacobian is very close
to singular for some values of the model parameters. In such cases, we need to find
an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
In our work, we prepared experimental prototypes of `nls` that incorporate
stabilization, but integration with all the features of `nls()` has proved difficult.

@@

```{r SESingGrad}
reala <- -3; realb <- 5; realc <- 0.5; realr <- 0.7; realm <- 1 # underlying parameters
x <- 1:11 # x values; 11 data points
y <- reala + realb * realr^(x - realm) + realc * x # linear + exponential function
testdat <- data.frame(x, y) # save in a data frame
strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
jform <- y ~ a + b * r^(x - m) + c * x # Formula
library(nlsr)
linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
linexp2 # Note singular values of Jacobian in rightmost column
# We get an error with nls()
linexp <- try(nls(jform, data = testdat, start = strt, trace = F))
```

@@

The typical default action, which can be seen by using the command `getOption("na.action")`
is `na.omit`. This option essentially omits from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much of the same for computations, but keeps the rows with NA elements so 
that predictions are in the correct row position. We recommend that workers actually test 
output to verify the behaviour is as wanted.

@@

`model` is an argument to the `nls()` function, which is documented as:

@@

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do NOT like this approach, but it is "the R way". R allows users to leave many 
objects in the default (.GlobalEnv) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the path, 
when R is launched, will be loaded. The overwhelming proportion of R users in our
acquaintance avoid saving the workspace because of the danger of lurking data and
functions which may cause unwanted results.

@@

Other concerns are revealed by the example below. Here, the `SSlogis` selfStart function 
is used to generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.

The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
<!-- ``` -->
<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
<!--     selfStart( -->
<!--         function(input, Asym, xmid, scal) -->
<!--         { -->
<!--               .expr1 <- xmid - input -->
<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
<!--               .expr4 <- 1 + .expr3 -->
<!--               .value <- Asym/.expr4 -->
<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
<!--               { -->
<!--             		  .expr10 <- .expr4^2 -->
<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
<!--                   .grad[, "Asym"] <- 1/.expr4 -->
<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
<!-- 		              .grad[, "scal"] <- xm * .e2 -->
<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
<!--                   attr(.value, "gradient") <- .grad -->
<!--               } -->
<!--               .value -->
<!--         }, -->
<!--         initial = function(mCall, data, LHS, ...) { -->
<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
<!--               if(nrow(xy) < 4) { -->
<!--                   stop("too few distinct input values to fit a logistic model") -->
<!--               } -->
<!--               z <- xy[["y"]] -->
<!--               ## transform to proportion, i.e. in (0,1) : -->
<!--               rng <- range(z); dz <- diff(rng) -->
<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
<!--               aux <- coef(lm(x ~ z, xy)) -->
<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
<!--                                data = xy, -->
<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
<!--                                algorithm = "plinear", ...)) -->
<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
<!--                        mCall[c("Asym", "xmid", "scal")]) -->
<!--         }, -->
<!--         parameters = c("Asym", "xmid", "scal")) -->
<!-- ``` -->
This R function includes analytic expressions for the Jacobian ("gradient").
These could be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
could be provided quite simply instead of all 1's.

@@

>A list of

>`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.

>`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.

>`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.

>`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.

>`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.

>`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.

>`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.

>`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.

>`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.

>`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.

> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.

@@

The `data` return element is an R symbol. To actually access the data from this 
element, we need to use the syntax:

@@

However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:

@@

>`coefficients`    A named vector giving the parameter values at the supposed solution.

>`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.

>`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.

>`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
                parameters) at the \newline
$~~~~~~~~~~~~~~~~~~~~$ returned parameters.

>`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.

>`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.

@@

```{r tetrarun}
time <- c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
conc <- c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
NLSdata <- data.frame(time,conc)
NLSstart <- c(lrc1 = -2, lrc2 = 0.25, A1 = 150, A2 = 50) # a starting vector (named!)
NLSformula <- conc ~ A1 * exp(-exp(lrc1) * time) + A2 * exp(-exp(lrc2) * time)
tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
print(tryit)
```

@@

One obvious case is that nonlinear modeling tools are a superset of linear ones.
Yet the explicit model `y ~ a * x + b` does not work with the linear modeling 
function `lm()`, which requires this model to be specified as `y ~ x`.

Within `nls()`, consider the 
following FOUR different calling sequences for the same problem, though an intuitive choice,
labelled fm2a, will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do
not foresee a resolution for this issue. In the example, we have NOT evaluated
the commands to save space.

@@

## using conditional linearity - leave out the Asym parameter
fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
                 data = DNase1, start = list(xmid = 0, scal = 1),
                 algorithm = "plinear")
summary(fm2)

## without conditional linearity
fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
                 data = DNase1,
                 start = list(Asym = 3, xmid = 0, scal = 1))
summary(fm3)

## using Port's nl2sol algorithm
fm4 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
                 algorithm = "port"))
summary(fm4)

## using conditional linearity AND Asym does NOT work
fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
                 algorithm = "plinear", trace = TRUE))
summary(fm2a)
```

@@

For `nls()` the concern is demonstrated by a short email query from John Nash
to Doug Bates and his reply.
This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on 
how difficult it is to develop code in this subject area and to keep it maintainable. 
We have experienced similar loss of understanding for some of our own codes.

@@

> The `algorithm = "port"` code appears unfinished, and does not even check that the starting 
  value is within the bounds. Use with caution, especially where bounds are supplied.

@@

```{r chkbound}
# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
wf <- data.frame(y = weed, tt = tt)
st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
require(minpack.lm)
require(nlsr)
# Hobbs scaled problem with bounds, formula specification
anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
cat("Using nlsr::nlxb():")
print(anlxb1)
## nlsLM seems NOT to work properly with bounds
anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
cat("\n"); cat("using minpack.lm::nlsLM():")
print(anlsLM1b)
anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", lower = c(0, 0, 0), upper = c(2, 6, 3))
cat("Using nls with 'port':")
print(anlsb)
```

@@

Bounds on parameters raise some interesting and difficult questions about how uncertainty
in parameter estimates should be computed or reported. That is, the traditional "standard
errors" are generally taken to imply symmetric intervals about the point estimate in which
the parameter may be expected to be found with some probability under certain assumptions.
Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.

@@

We want:

@@

- to streamline the overall `nls()` infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as new features we or others
  would like to add.

@@

While not consistent with the `lm()` specification of linear models, 
nonlinear models need all parameters specified. 
That is, the model formula should be written down as it would be computed.
Thus a linear model (which is valid input for nonlinear estimatin) should
be specified `y ~ a * x + b` rather than `y ~ x`. 
Moreover, we want to avoid the partially linear parameters appearing 
in the result object as `.lin1`, `.lin2`, etc., where the actual position 
of parameters in the model is not explicit.

@@

A further complication of model specification in R is that some formulas require
the user to surround a term with `I()` to inhibit interpretation of arithmetic 
operators. We would prefer that formulas be usable without this complication.

@@

Indexed models clearly have a place in some areas of research. 
<!-- We need to be able to process models, as the example above, where the -->
<!-- model formula is  -->
We need to be able to process models such as 
`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)` where `Strip` is the index. Data
`Length` and `Conc` are available for all observations and parameter `th` applies
to every fitted point, but `a[Strip]` and `b[Strip]` depend on the index, which
could be an experimental run number, or some other categorization.
